---
title: "Getting started with taupatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with taupatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 120
)

# Everything below runs on synthetic data, so it needs no network and no
# Copernicus credentials. It does need the default model engine, which is a
# Suggests rather than a hard dependency.
runnable <- requireNamespace("ranger", quietly = TRUE) &&
  requireNamespace("datamatch", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = runnable)
```

```{r not-runnable, eval = !runnable, echo = FALSE, results = "asis"}
cat("> **The code below is not evaluated in this build.** It needs `ranger`",
    "and `datamatch`, which are suggested rather than required. Install them",
    "and the vignette runs end to end in a couple of seconds.\n")
```

```{r library}
library(taupatch)
```

## What a run does

taupatch answers one question, monthly, over an area: **where are the patches?**
A patch is a station whose abundance of some zooplankton species exceeds a
threshold — the top decile, say, or a stated number of animals per square metre.
Everything else is a non-patch.

That turns an abundance survey into a presence/absence problem, which a
classifier can be fitted to and a fitted classifier can be projected onto a grid.
Five stages, in order:

1. **Load** the station data and label each station patch or non-patch.
2. **Fetch** environmental covariates for the study area and match them to the
   stations.
3. **Fit** a model of patch probability on those covariates.
4. **Evaluate** it against held-out folds.
5. **Project** it onto the covariate grid, one map per month.

One config file describes all five. Two runs differ by a file rather than by
edited code, which is the whole design.

## A run you can do right now

The package ships a synthetic dataset and a config that uses it, so the pipeline
can be exercised before you have Copernicus credentials or a station database.
The covariate source is `mock`: nothing is downloaded.

```{r config}
config <- load_config(system.file("configs/mock_test.yaml", package = "taupatch"))

# Write into the session's temp directory rather than the installed package.
config$paths$zoop_file <- file.path(tempdir(), "mock_stations.csv")
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette")

generate_mock_zoop_data(config)
```

`load_config()` has already done more than read the file. It applied the
defaults, resolved every relative path against the config's own location, and
checked what is cheap to check: that the species resolves, the thresholds are
well formed, the covariate names exist, and the columns it declares are actually
in the CSV. A mistyped covariate is an error here, at no cost, rather than five
minutes into a Copernicus download.

Now run it.

```{r run, message = FALSE}
result <- run_taupatch(config)
```

That is the whole pipeline. On real data the same call takes as long as the
downloads do; here it takes a couple of seconds.

## Look at the stations first

Before reading any model output, look at what it was fitted on. This is the step
that catches a study area drawn wider than the survey, or an abundance field that
is mostly zeros.

```{r stations}
station_summary(result$data)
```

Median abundance and the 90th percentile are the two numbers to check against
each other. The config asks for a 90th-percentile threshold, so that second
number *is* the patch threshold, and roughly a tenth of stations are patches by
construction. Hold on to that — it decides how the evaluation has to be read.

```{r station-map, fig.height = 5}
plot_station_map(result$data)
```

## What came back

`run_taupatch()` returns the fitted model, the projections, and the data behind
both.

```{r result-names}
names(result)
names(result$model)
```

### Which covariates mattered

```{r importance}
result$model$importance
```

Importance here is the drop in ROC AUC when a predictor is shuffled, averaged
over several shuffles. It is computed the same way whichever model type was
fitted, which is what makes a random forest's ranking comparable with a GAM's —
engine-reported importances are not, since ranger's permutation drop and
xgboost's split gain are different quantities on different scales.

It says a predictor matters. It does not say what it *does*.

### What each covariate does

```{r partial, fig.height = 3.2}
effects <- partial_effects(result$model$workflow,
                           result$model$model_data,
                           result$model$predictors)
plot_partial_effects(effects)
```

Each panel sweeps one predictor across its range with the others held at the
values they actually take, and plots the mean predicted patch probability. The
mock data plants a latitudinal gradient and a seasonal cycle into both abundance
and the covariates, so these curves should bend — a flat panel on real data means
a predictor that is carrying nothing.

## Reading the evaluation

This is the part worth slowing down for, because the obvious reading of it is
wrong.

```{r evals}
result$model$evaluation[, c("metric", "threshold", "value", "std_err",
                            "lower", "upper")]
```

Every threshold-dependent metric appears twice, at two cutoffs, and the table
says which is which. That is deliberate.

The two uncertainty columns answer different questions, which is why blanks in
one do not mean missing data. `std_err` is the spread across cross-validation
folds — only the metrics averaged per fold have one, and `pr_auc`, `tss` and
`precision` come from the pooled held-out predictions instead. `lower` and
`upper` bootstrap those pooled predictions, so every row has an interval: one
column asks how much the number moves when the model is refitted, the other how
much it moves if the survey had sampled different stations.

```{r cutoff}
result$model$classification_threshold
result$model$classification_threshold_interval
```

That cutoff is estimated from the same held-out predictions it then scores, so
it moves too — the interval above is how far, and each bootstrap resample
re-derives its own rather than being handed this one. A wide range there means a
binarised map depends on which stations happened to be sampled.

**At the default 0.5 cutoff the model looks much worse than it is.** With only a
tenth of stations patches, a classifier that is genuinely good at ranking will
still put almost every station below 0.5, so sensitivity reads terribly. Compare
the two `sens` rows above: the same model, the same predictions, a different
number. Use `classification_threshold` when you binarise a projection — it is
also written to `threshold.yaml` next to the outputs.

**ROC AUC flatters an imbalanced problem.** Its false-positive rate has the large
non-patch class in the denominator, so it stays high while precision does not.
`pr_auc` is the honest companion, and the gap between the two is the thing to
watch (Saito & Rehmsmeier 2015; Sofaer et al. 2019).

The abundance threshold itself — the number of animals a patch starts at — is
separate from the probability cutoff, and is recorded too:

```{r threshold}
result$model$threshold
```

A percentile run can therefore be reproduced later as an absolute one.

## The maps

One projection per configured month and year, written as a GeoTIFF and a PNG,
and pooled into a single CSV carrying coordinates and dates.

```{r projections}
result$projections[, c("year", "month", "n_cells", "n_grid", "resolution")]
```

`n_cells` against `n_grid` is the gap report: how many cells of the covariate
grid actually got a prediction. When they disagree, a covariate was missing there
— cloud in the satellite chlorophyll, or a neighbourhood-derived covariate that
is undefined on the border of the study area. That is what explains a patchy map,
and the run log names the covariate responsible.

The projection is onto the covariate grid at its own resolution, not back onto
the station locations, so the map is a surface rather than a scatter of points.

```{r map, fig.height = 5}
suitability <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)

library(ggplot2)
ggplot(subset(suitability, year == 2018 & month == 7),
       aes(lon, lat, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", limits = c(0, 1), name = "P(patch)") +
  coord_quickmap() +
  labs(title = "cfin patch probability, July 2018", x = NULL, y = NULL) +
  theme_bw()
```

## How far to trust the map

Everything above is a point estimate. `projection.uncertainty` adds two layers
that say where not to believe it — off by default, because on a real grid the
extra prediction passes cost real time.

```{r uncertainty, message = FALSE}
config$projection$uncertainty <- TRUE
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette_unc")

result <- run_taupatch(config)

suitability <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)
names(suitability)
```

The models behind the interval are the ones cross-validation already fitted and
would otherwise have thrown away, so the ensemble itself is free.

```{r ensemble}
length(result$model$ensemble)
```

Two layers, two different questions. `suitability_sd` and the interval say how
much the surface moves when the model is refitted on resampled data.
`novelty` says whether the cell is somewhere the model has ever been:

```{r novelty}
summary(suitability$novelty)

# Which covariate puts a cell outside the training range, where one is.
table(suitability$novel_variable[suitability$novelty < 0])
```

Novelty runs to 100 at the median of the training data and **goes negative
outside its range** — those cells are extrapolation, and the model has no
evidence for what it claims there.

Read both, and do not treat a narrow interval as reassurance on its own. Every
ensemble member was trained on the same data, so they can walk off the end of it
together and agree the whole way. That combination — confident and novel — is
the one to look for:

```{r disagree}
extrapolated <- suitability$novelty < 0
data.frame(
  cells = c(sum(!extrapolated), sum(extrapolated)),
  median_sd = round(c(median(suitability$suitability_sd[!extrapolated]),
                      median(suitability$suitability_sd[extrapolated])), 4),
  row.names = c("inside training range", "outside training range")
)
```

Each month gets a two-panel figure in `plots/`, and the layers go into the
GeoTIFF beside the mean rather than into a second file:

```{r layers}
names(terra::rast(result$projections$geotiff[1]))
```

## What is on disk

```{r outputs}
list.files(config$paths$output_dir, recursive = TRUE)[1:20]
```

`model.rds` is the fitted workflow, so a projection can be redone without
refitting. `diagnostics/cv_predictions.csv` holds the held-out predictions, so
any metric not in `evals.csv` can be computed without refitting either.

## Which covariates are earning their place

Importance ranks the covariates a model has. It cannot say whether a covariate
is carrying anything the others were not already carrying — a predictor can rank
third and be entirely redundant. `jackknife_covariates()` answers that by
refitting without each one, over the same folds, and seeing how much worse the
model ranks stations.

It needs no config block; the defaults are used when there is none.

```{r jackknife, message = FALSE}
jk <- jackknife_covariates(result$data, config)

jk[, c("variable", "score_full", "score_without", "score_only", "contribution")]
```

Two halves, two questions. **`score_without`** is low when the covariate carries
something no other covariate has — its *unique* contribution, which is what
`contribution` measures. **`score_only`** is the model on that covariate alone,
so it is high when the covariate carries a lot whether or not anything else
carries it too.

Read the two columns against each other and the mock run says something the
importance table above could not. `jday` scores well on its own — the synthetic
data has a real seasonal cycle in it — and contributes nothing on top of the
others, which have absorbed the same seasonality. That is information which is
real *and* duplicated, and it is a very different situation from information
that is not there. Ranking on `contribution` alone would treat the two
identically.

Some contributions come out negative. That means the model scored *better* on
average without the covariate, which on this many stations is noise rather than
a finding — and it is exactly the case the test exists to keep you from
over-reading.

```{r jackknife-test}
jk[, c("variable", "contribution", "statistic", "p_value", "p_adjusted",
       "significant")]
```

Nothing here is significant, and that is the right answer rather than a broken
test. Three covariates over a few hundred synthetic stations, scored on five
folds, is not enough evidence to establish that any one of them is load-bearing.
A test that returned confident answers from this much data would be the one to
distrust.

`p_value` is one-sided on the per-fold differences, with the variance inflated
to account for the folds sharing most of their training rows. At five folds that
inflation is a factor of 1.5 on the standard error, so every statistic here is
two-thirds of what a naive paired *t*-test would have reported. It makes no
difference to the verdict on this run — these differences are nowhere near the
boundary either way — but it is the difference between a real finding and a
false one when a covariate lands close to it, which on real data is where the
interesting ones land. `p_adjusted` then accounts for having asked once per
covariate. A GLM or a GAM would also fill in `parametric_p` with a likelihood
ratio test; a forest has no likelihood, so it is `NA` here.

**Nothing is dropped.** That is the default, and it is deliberate:

```{r jackknife-drop}
jackknife_settings(config)   # NULL: this config has no jackknife block

# What dropping *would* have removed, under the settings the test ran with.
jackknife_dropped(jk)
```

Only one name comes back even though all three failed, because
`min_predictors` floors the model at two covariates and the ones handed back are
the strongest of those that failed.

A covariate that fails this test is one the *others already account for on these
stations*, which is a statement about collinearity in this sample at least as
much as about ecology. Depth and surface temperature carry much of the same
information on a shelf, and the test will call either one redundant depending on
which the model reached for first. Turning `covariates.jackknife.drop` on tells
the pipeline to act on it anyway, and `keep` protects a covariate that is in the
model because the study is about it:

```{r jackknife-config, eval = FALSE}
config$covariates$jackknife <- list(drop = TRUE, keep = "jday", workers = 4)
result <- run_taupatch(config)

result$jackknife                      # the table, also written to the run
result$config$covariates$exclude      # what came out
```

## Fitting all the models at once

Rather than choosing an algorithm, fit several and combine them. Set
`model.type` to `ensemble` and the pipeline does the rest — everything after the
fit works the same, so nothing else in the config changes.

Four types are available; two are used here because `xgboost` and `mgcv` are
suggested rather than required.

```{r ensemble-fit, message = FALSE, warning = FALSE}
config$model$ensemble <- list(types = c("rf", "glm"), rule = "weighted_mean")
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette_ens")

result <- run_taupatch(config)
result$model
```

Each member's score, and the weight it earned, is the first thing to read. A
table showing one member near 1.0 and the rest near zero is a single model with
extra steps.

```{r ensemble-summary}
result$model$summary
```

The ensemble's evaluation is **its own**, not the average of its members'. Every
member was fitted on the same folds from the same seed, so their held-out
predictions line up row for row, and the combined out-of-fold predictions go
through the same evaluation as a single model's:

```{r ensemble-eval}
result$model$evaluation[1:2, c("metric", "value", "lower", "upper")]

# Each member's own, for comparison.
vapply(result$model$members,
       function(m) m$evaluation$value[m$evaluation$metric == "roc_auc" &
                                        is.na(m$evaluation$threshold)],
       numeric(1))
```

That distinction matters. Combining members that make *different* mistakes beats
all of them; combining members that make the same mistakes does not, and only a
cross-validated ensemble prediction can tell those apart.

Importance is weighted across members, with each member's own kept beside it — a
predictor the forest leans on and the GLM ignores is a fact about the shape of
the relationship, and the average is the one number that hides it.

```{r ensemble-importance}
result$model$importance
```

Every combination rule is computed and written, so a committee map can be read
off the same file without refitting. `rule` only picks which one becomes the
`suitability` layer:

```{r ensemble-layers}
names(terra::rast(result$projections$geotiff[1]))
```

`algorithm_sd` is the new one, and it is **not** the same as `suitability_sd`
from the section above. That one is a single algorithm refitted on resampled
stations; this one is different algorithms looking at the same stations and
disagreeing. Both can be on at once, and a cell can be quiet on one and loud on
the other:

```{r ensemble-spread}
ens <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)

summary(ens[, c("suitability", "algorithm_sd", "suitability_sd")])
```

## Moving to real data

Three things change, and nothing else does.

**The station file.** Point `paths.zoop_file` at your own CSV. If you have the
raw NOAA EcoMon export rather than a formatted database, reshape it first —
`format_zoop_data()` splits the date, renames the coordinate columns, strips the
units off each taxon column and divides the abundances accordingly, and carries
everything else through untouched.

```{r format, eval = FALSE}
format_zoop_data("raw_ecomon.csv", write_to = "data/zooplankton.csv")
zoop_taxa("raw_ecomon.csv")   # every taxon the file carries
```

**The covariate source.** `covariates.source: copernicus` instead of `mock`, and
the [Copernicus Marine
Toolbox](https://help.marine.copernicus.eu/en/collections/4060068-copernicus-marine-toolbox)
installed and logged in. The covariate names stay the same:

```{r catalog}
covariate_info()[1:8, c("name", "label", "units", "spatial")]
```

**The window and the area.** `dates`, `projection`, and `study_area.bbox`. Draw
the bounding box wider than your stations if you use gradients or fronts, which
are undefined on the edge of the grid.

Write the new config rather than hand-editing YAML — `generate_config()`
validates as it writes, so a mistyped covariate is an error at that call:

```{r generate, eval = FALSE}
generate_config("cfin_gom",
                zoop_file = "data/zooplankton.csv",
                selected = c("SST", "SSS", "CHL"),
                bathymetry = "DEPTH",
                transform = list(log1p = c("CHL", "DEPTH")))
```

## Doing it without writing code

The same pipeline is a GUI. Every config field is a control, the run streams its
progress, and **Download config** turns whatever you clicked into a YAML file you
can re-run or hand to someone else.

```{r app, eval = FALSE}
run_taupatch_app()
```

## Where to go next

The README covers each config block in depth — thresholds, derived covariates,
transformations, resampling before the join, the four model types and the
diagnostics each one contributes. In R:

- `?load_config`, `?generate_config` — the config
- `?covariate_info`, `?derivoce_covariates` — what can be a predictor
- `?covariate_transforms` — and what can be done to one
- `?model_types` — the four models, and what each is good for
- `?partial_effects`, `?gam_smooth_terms` — reading a fitted model
- `?jackknife_covariates`, `?jackknife_settings` — testing which covariates earn
  their place
- `?fit_patch_ensemble`, `?ensemble_settings`, `?ensemble_rules` — combining
  several model types

## Citing a run

taupatch is mostly plumbing between other people's data and other people's
methods. A run reaches Copernicus Marine products, NOAA terrain, and station data
none of which belong to this package, and fits models and metrics that are each
somebody's paper. `citation("taupatch")` gives this package and the paper it
implements; the [References section of the
README](https://github.com/chross22/taupatch#references) groups everything else
so you can pick out the subset a particular run actually used. Each function's
own help page carries the references relevant to it.
