---
title: "Getting started with taupatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with taupatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 120
)

# Everything below runs on synthetic data, so it needs no network and no
# Copernicus credentials. It does need the default model engine, which is a
# Suggests rather than a hard dependency.
runnable <- requireNamespace("ranger", quietly = TRUE) &&
  requireNamespace("datamatch", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = runnable)
```

```{r not-runnable, eval = !runnable, echo = FALSE, results = "asis"}
cat("> **The code below is not evaluated in this build.** It needs `ranger`",
    "and `datamatch`, which are suggested rather than required. Install them",
    "and the vignette runs end to end in a couple of seconds.\n")
```

```{r library}
library(taupatch)
```

## What a run does

taupatch answers one question, monthly, over an area: **where are the patches?**
A patch is a station whose abundance of some zooplankton species exceeds a
threshold — the top decile, say, or a stated number of animals per square metre.
Everything else is a non-patch.

That turns an abundance survey into a presence/absence problem, which a
classifier can be fitted to and a fitted classifier can be projected onto a grid.
Five stages, in order:

1. **Load** the station data and label each station patch or non-patch.
2. **Fetch** environmental covariates for the study area and match them to the
   stations.
3. **Fit** a model of patch probability on those covariates.
4. **Evaluate** it against held-out folds.
5. **Project** it onto the covariate grid, one map per month.

One config file describes all five. Two runs differ by a file rather than by
edited code, which is the whole design.

## A run you can do right now

The package ships a synthetic dataset and a config that uses it, so the pipeline
can be exercised before you have Copernicus credentials or a station database.
The covariate source is `mock`: nothing is downloaded.

```{r config}
config <- load_config(system.file("configs/mock_test.yaml", package = "taupatch"))

# Write into the session's temp directory rather than the installed package.
config$paths$zoop_file <- file.path(tempdir(), "mock_stations.csv")
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette")

generate_mock_zoop_data(config)
```

`load_config()` has already done more than read the file. It applied the
defaults, resolved every relative path against the config's own location, and
checked what is cheap to check: that the species resolves, the thresholds are
well formed, the covariate names exist, and the columns it declares are actually
in the CSV. A mistyped covariate is an error here, at no cost, rather than five
minutes into a Copernicus download.

Now run it.

```{r run, message = FALSE}
result <- run_taupatch(config)
```

That is the whole pipeline. On real data the same call takes as long as the
downloads do; here it takes a couple of seconds.

## Look at the stations first

Before reading any model output, look at what it was fitted on. This is the step
that catches a study area drawn wider than the survey, or an abundance field that
is mostly zeros.

```{r stations}
station_summary(result$data)
```

Median abundance and the 90th percentile are the two numbers to check against
each other. The config asks for a 90th-percentile threshold, so that second
number *is* the patch threshold, and roughly a tenth of stations are patches by
construction. Hold on to that — it decides how the evaluation has to be read.

```{r station-map, fig.height = 5}
plot_station_map(result$data)
```

## What came back

`run_taupatch()` returns the fitted model, the projections, and the data behind
both.

```{r result-names}
names(result)
names(result$model)
```

### Which covariates mattered

```{r importance}
result$model$importance
```

Importance here is the drop in ROC AUC when a predictor is shuffled, averaged
over several shuffles. It is computed the same way whichever model type was
fitted, which is what makes a random forest's ranking comparable with a GAM's —
engine-reported importances are not, since ranger's permutation drop and
xgboost's split gain are different quantities on different scales.

It says a predictor matters. It does not say what it *does*.

### What each covariate does

```{r partial, fig.height = 3.2}
effects <- partial_effects(result$model$workflow,
                           result$model$model_data,
                           result$model$predictors)
plot_partial_effects(effects)
```

Each panel sweeps one predictor across its range with the others held at the
values they actually take, and plots the mean predicted patch probability. The
mock data plants a latitudinal gradient and a seasonal cycle into both abundance
and the covariates, so these curves should bend — a flat panel on real data means
a predictor that is carrying nothing.

## Reading the evaluation

This is the part worth slowing down for, because the obvious reading of it is
wrong.

```{r evals}
result$model$evaluation[, c("metric", "threshold", "value")]
```

Every threshold-dependent metric appears twice, at two cutoffs, and the table
says which is which. That is deliberate.

```{r cutoff}
result$model$classification_threshold
```

**At the default 0.5 cutoff the model looks much worse than it is.** With only a
tenth of stations patches, a classifier that is genuinely good at ranking will
still put almost every station below 0.5, so sensitivity reads terribly. Compare
the two `sens` rows above: the same model, the same predictions, a different
number. Use `classification_threshold` when you binarise a projection — it is
also written to `threshold.yaml` next to the outputs.

**ROC AUC flatters an imbalanced problem.** Its false-positive rate has the large
non-patch class in the denominator, so it stays high while precision does not.
`pr_auc` is the honest companion, and the gap between the two is the thing to
watch (Saito & Rehmsmeier 2015; Sofaer et al. 2019).

The abundance threshold itself — the number of animals a patch starts at — is
separate from the probability cutoff, and is recorded too:

```{r threshold}
result$model$threshold
```

A percentile run can therefore be reproduced later as an absolute one.

## The maps

One projection per configured month and year, written as a GeoTIFF and a PNG,
and pooled into a single CSV carrying coordinates and dates.

```{r projections}
result$projections[, c("year", "month", "n_cells", "n_grid", "resolution")]
```

`n_cells` against `n_grid` is the gap report: how many cells of the covariate
grid actually got a prediction. When they disagree, a covariate was missing there
— cloud in the satellite chlorophyll, or a neighbourhood-derived covariate that
is undefined on the border of the study area. That is what explains a patchy map,
and the run log names the covariate responsible.

The projection is onto the covariate grid at its own resolution, not back onto
the station locations, so the map is a surface rather than a scatter of points.

```{r map, fig.height = 5}
suitability <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)

library(ggplot2)
ggplot(subset(suitability, year == 2018 & month == 7),
       aes(lon, lat, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", limits = c(0, 1), name = "P(patch)") +
  coord_quickmap() +
  labs(title = "cfin patch probability, July 2018", x = NULL, y = NULL) +
  theme_bw()
```

## How far to trust the map

Everything above is a point estimate. `projection.uncertainty` adds two layers
that say where not to believe it — off by default, because on a real grid the
extra prediction passes cost real time.

```{r uncertainty, message = FALSE}
config$projection$uncertainty <- TRUE
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette_unc")

result <- run_taupatch(config)

suitability <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)
names(suitability)
```

The models behind the interval are the ones cross-validation already fitted and
would otherwise have thrown away, so the ensemble itself is free.

```{r ensemble}
length(result$model$ensemble)
```

Two layers, two different questions. `suitability_sd` and the interval say how
much the surface moves when the model is refitted on resampled data.
`novelty` says whether the cell is somewhere the model has ever been:

```{r novelty}
summary(suitability$novelty)

# Which covariate puts a cell outside the training range, where one is.
table(suitability$novel_variable[suitability$novelty < 0])
```

Novelty runs to 100 at the median of the training data and **goes negative
outside its range** — those cells are extrapolation, and the model has no
evidence for what it claims there.

Read both, and do not treat a narrow interval as reassurance on its own. Every
ensemble member was trained on the same data, so they can walk off the end of it
together and agree the whole way. That combination — confident and novel — is
the one to look for:

```{r disagree}
extrapolated <- suitability$novelty < 0
data.frame(
  cells = c(sum(!extrapolated), sum(extrapolated)),
  median_sd = round(c(median(suitability$suitability_sd[!extrapolated]),
                      median(suitability$suitability_sd[extrapolated])), 4),
  row.names = c("inside training range", "outside training range")
)
```

Each month gets a two-panel figure in `plots/`, and the layers go into the
GeoTIFF beside the mean rather than into a second file:

```{r layers}
names(terra::rast(result$projections$geotiff[1]))
```

## What is on disk

```{r outputs}
list.files(config$paths$output_dir, recursive = TRUE)[1:20]
```

`model.rds` is the fitted workflow, so a projection can be redone without
refitting. `diagnostics/cv_predictions.csv` holds the held-out predictions, so
any metric not in `evals.csv` can be computed without refitting either.

## Moving to real data

Three things change, and nothing else does.

**The station file.** Point `paths.zoop_file` at your own CSV. If you have the
raw NOAA EcoMon export rather than a formatted database, reshape it first —
`format_zoop_data()` splits the date, renames the coordinate columns, strips the
units off each taxon column and divides the abundances accordingly, and carries
everything else through untouched.

```{r format, eval = FALSE}
format_zoop_data("raw_ecomon.csv", write_to = "data/zooplankton.csv")
zoop_taxa("raw_ecomon.csv")   # every taxon the file carries
```

**The covariate source.** `covariates.source: copernicus` instead of `mock`, and
the [Copernicus Marine
Toolbox](https://help.marine.copernicus.eu/en/collections/4060068-copernicus-marine-toolbox)
installed and logged in. The covariate names stay the same:

```{r catalog}
covariate_info()[1:8, c("name", "label", "units", "spatial")]
```

**The window and the area.** `dates`, `projection`, and `study_area.bbox`. Draw
the bounding box wider than your stations if you use gradients or fronts, which
are undefined on the edge of the grid.

Write the new config rather than hand-editing YAML — `generate_config()`
validates as it writes, so a mistyped covariate is an error at that call:

```{r generate, eval = FALSE}
generate_config("cfin_gom",
                zoop_file = "data/zooplankton.csv",
                selected = c("SST", "SSS", "CHL"),
                bathymetry = "DEPTH",
                transform = list(log1p = c("CHL", "DEPTH")))
```

## Doing it without writing code

The same pipeline is a GUI. Every config field is a control, the run streams its
progress, and **Download config** turns whatever you clicked into a YAML file you
can re-run or hand to someone else.

```{r app, eval = FALSE}
run_taupatch_app()
```

## Where to go next

The README covers each config block in depth — thresholds, derived covariates,
transformations, resampling before the join, the four model types and the
diagnostics each one contributes. In R:

- `?load_config`, `?generate_config` — the config
- `?covariate_info`, `?derivoce_covariates` — what can be a predictor
- `?covariate_transforms` — and what can be done to one
- `?model_types` — the four models, and what each is good for
- `?partial_effects`, `?gam_smooth_terms` — reading a fitted model

## Citing a run

taupatch is mostly plumbing between other people's data and other people's
methods. A run reaches Copernicus Marine products, NOAA terrain, and station data
none of which belong to this package, and fits models and metrics that are each
somebody's paper. `citation("taupatch")` gives this package and the paper it
implements; the [References section of the
README](https://github.com/chross22/taupatch#references) groups everything else
so you can pick out the subset a particular run actually used. Each function's
own help page carries the references relevant to it.
