---
title: "Getting started with taupatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with taupatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 120
)

# Everything below runs on synthetic data, so it needs no network and no
# Copernicus credentials. It does need the default model engine, which is a
# Suggests rather than a hard dependency.
runnable <- requireNamespace("ranger", quietly = TRUE) &&
  requireNamespace("datamatch", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(eval = runnable)
```

```{r not-runnable, eval = !runnable, echo = FALSE, results = "asis"}
cat("> **The code below is not evaluated in this build.** It needs `ranger`",
    "and `datamatch`, which are suggested rather than required. Install them",
    "and the vignette runs end to end in a couple of seconds.\n")
```

```{r library}
library(taupatch)
```

## What a run does

taupatch answers one question, monthly, over an area: **where are the patches?**
A patch is a station whose abundance of some zooplankton species exceeds a
threshold — the top decile, say, or a stated number of animals per square metre.
Everything else is a non-patch.

That turns an abundance survey into a presence/absence problem, which a
classifier can be fitted to and a fitted classifier can be projected onto a grid.
Five stages, in order:

1. **Load** the station data and label each station patch or non-patch.
2. **Fetch** environmental covariates for the study area and match them to the
   stations.
3. **Fit** a model of patch probability on those covariates.
4. **Evaluate** it against held-out folds.
5. **Project** it onto the covariate grid, one map per month.

One config file describes all five. Two runs differ by a file rather than by
edited code, which is the whole design.

## A run you can do right now

The package ships a synthetic dataset and a config that uses it, so the pipeline
can be exercised before you have Copernicus credentials or a station database.
The covariate source is `mock`: nothing is downloaded.

```{r config}
config <- load_config(system.file("configs/mock_test.yaml", package = "taupatch"))

# Write into the session's temp directory rather than the installed package.
config$paths$zoop_file <- file.path(tempdir(), "mock_stations.csv")
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette")

generate_mock_zoop_data(config)
```

`load_config()` has already done more than read the file. It applied the
defaults, resolved every relative path against the config's own location, and
checked what is cheap to check: that the species resolves, the thresholds are
well formed, the covariate names exist, and the columns it declares are actually
in the CSV. A mistyped covariate is an error here, at no cost, rather than five
minutes into a Copernicus download.

Now run it.

```{r run, message = FALSE}
result <- run_taupatch(config)
```

That is the whole pipeline. On real data the same call takes as long as the
downloads do; here it takes a couple of seconds.

## Look at the stations first

Before reading any model output, look at what it was fitted on. This is the step
that catches a study area drawn wider than the survey, or an abundance field that
is mostly zeros.

```{r stations}
station_summary(result$data)
```

Median abundance and the 90th percentile are the two numbers to check against
each other. The config asks for a 90th-percentile threshold, so that second
number *is* the patch threshold, and roughly a tenth of stations are patches by
construction. Hold on to that — it decides how the evaluation has to be read.

```{r station-map, fig.height = 5}
plot_station_map(result$data)
```

## What came back

`run_taupatch()` returns the fitted model, the projections, and the data behind
both.

```{r result-names}
names(result)
names(result$model)
```

### Which covariates mattered

```{r importance}
result$model$importance
```

Importance here is the drop in ROC AUC when a predictor is shuffled, averaged
over several shuffles. It is computed the same way whichever model type was
fitted, which is what makes a random forest's ranking comparable with a GAM's —
engine-reported importances are not, since ranger's permutation drop and
xgboost's split gain are different quantities on different scales.

It says a predictor matters. It does not say what it *does*.

### What each covariate does

```{r partial, fig.height = 3.2}
effects <- partial_effects(result$model$workflow,
                           result$model$model_data,
                           result$model$predictors)
plot_partial_effects(effects)
```

Each panel sweeps one predictor across its range with the others held at the
values they actually take, and plots the mean predicted patch probability. The
mock data plants a latitudinal gradient and a seasonal cycle into both abundance
and the covariates, so these curves should bend — a flat panel on real data means
a predictor that is carrying nothing.

## Reading the evaluation

This is the part worth slowing down for, because the obvious reading of it is
wrong.

```{r evals}
result$model$evaluation[, c("metric", "threshold", "value", "std_err",
                            "lower", "upper")]
```

Every threshold-dependent metric appears twice, at two cutoffs, and the table
says which is which. That is deliberate.

The two uncertainty columns answer different questions, which is why blanks in
one do not mean missing data. `std_err` is the spread across cross-validation
folds — only the metrics averaged per fold have one, and `pr_auc`, `tss` and
`precision` come from the pooled held-out predictions instead. `lower` and
`upper` bootstrap those pooled predictions, so every row has an interval: one
column asks how much the number moves when the model is refitted, the other how
much it moves if the survey had sampled different stations.

```{r cutoff}
result$model$classification_threshold
result$model$classification_threshold_interval
```

That cutoff is estimated from the same held-out predictions it then scores, so
it moves too — the interval above is how far, and each bootstrap resample
re-derives its own rather than being handed this one. A wide range there means a
binarised map depends on which stations happened to be sampled.

**At the default 0.5 cutoff the model looks much worse than it is.** With only a
tenth of stations patches, a classifier that is genuinely good at ranking will
still put almost every station below 0.5, so sensitivity reads terribly. Compare
the two `sens` rows above: the same model, the same predictions, a different
number. Use `classification_threshold` when you binarise a projection — it is
also written to `threshold.yaml` next to the outputs.

**ROC AUC flatters an imbalanced problem.** Its false-positive rate has the large
non-patch class in the denominator, so it stays high while precision does not.
`pr_auc` is the honest companion, and the gap between the two is the thing to
watch (Saito & Rehmsmeier 2015; Sofaer et al. 2019).

The abundance threshold itself — the number of animals a patch starts at — is
separate from the probability cutoff, and is recorded too:

```{r threshold}
result$model$threshold
```

A percentile run can therefore be reproduced later as an absolute one.

## The maps

One projection per configured month and year, written as a GeoTIFF and a PNG,
and pooled into a single CSV carrying coordinates and dates.

```{r projections}
result$projections[, c("year", "month", "n_cells", "n_grid", "resolution")]
```

`n_cells` against `n_grid` is the gap report: how many cells of the covariate
grid actually got a prediction. When they disagree, a covariate was missing there
— cloud in the satellite chlorophyll, or a neighbourhood-derived covariate that
is undefined on the border of the study area. That is what explains a patchy map,
and the run log names the covariate responsible.

The projection is onto the covariate grid at its own resolution, not back onto
the station locations, so the map is a surface rather than a scatter of points.

```{r map, fig.height = 5}
suitability <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)

library(ggplot2)
ggplot(subset(suitability, year == 2018 & month == 7),
       aes(lon, lat, fill = suitability)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", limits = c(0, 1), name = "P(patch)") +
  coord_quickmap() +
  labs(title = "cfin patch probability, July 2018", x = NULL, y = NULL) +
  theme_bw()
```

## How far to trust the map

Everything above is a point estimate. `projection.uncertainty` adds two layers
that say where not to believe it — off by default, because on a real grid the
extra prediction passes cost real time.

```{r uncertainty, message = FALSE}
config$projection$uncertainty <- TRUE
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette_unc")

result <- run_taupatch(config)

suitability <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)
names(suitability)
```

The models behind the interval are the ones cross-validation already fitted and
would otherwise have thrown away, so the ensemble itself is free.

```{r ensemble}
length(result$model$ensemble)
```

Two layers, two different questions. `suitability_sd` and the interval say how
much the surface moves when the model is refitted on resampled data.
`novelty` says whether the cell is somewhere the model has ever been:

```{r novelty}
summary(suitability$novelty)

# Which covariate puts a cell outside the training range, where one is.
table(suitability$novel_variable[suitability$novelty < 0])
```

Novelty runs to 100 at the median of the training data and **goes negative
outside its range** — those cells are extrapolation, and the model has no
evidence for what it claims there.

Read both, and do not treat a narrow interval as reassurance on its own. Every
ensemble member was trained on the same data, so they can walk off the end of it
together and agree the whole way. That combination — confident and novel — is
the one to look for:

```{r disagree}
extrapolated <- suitability$novelty < 0
data.frame(
  cells = c(sum(!extrapolated), sum(extrapolated)),
  median_sd = round(c(median(suitability$suitability_sd[!extrapolated]),
                      median(suitability$suitability_sd[extrapolated])), 4),
  row.names = c("inside training range", "outside training range")
)
```

Each month gets a two-panel figure in `plots/`, and the layers go into the
GeoTIFF beside the mean rather than into a second file:

```{r layers}
names(terra::rast(result$projections$geotiff[1]))
```

## What is on disk

```{r outputs}
list.files(config$paths$output_dir, recursive = TRUE)[1:20]
```

`model.rds` is the fitted workflow, so a projection can be redone without
refitting. `diagnostics/cv_predictions.csv` holds the held-out predictions, so
any metric not in `evals.csv` can be computed without refitting either.

## Which covariates are earning their place

Importance ranks the covariates a model has. It cannot say whether a covariate
is carrying anything the others were not already carrying — a predictor can rank
third and be entirely redundant. `jackknife_covariates()` answers that by
refitting without each one, over the same folds, and seeing how much worse the
model ranks stations.

It needs no config block; the defaults are used when there is none.

```{r jackknife, message = FALSE}
jk <- jackknife_covariates(result$data, config)

jk[, c("variable", "score_full", "score_without", "score_only", "contribution")]
```

Two halves, two questions. **`score_without`** is low when the covariate carries
something no other covariate has — its *unique* contribution, which is what
`contribution` measures. **`score_only`** is the model on that covariate alone,
so it is high when the covariate carries a lot whether or not anything else
carries it too.

Read the two columns against each other and the mock run says something the
importance table above could not. `jday` scores well on its own — the synthetic
data has a real seasonal cycle in it — and contributes nothing on top of the
others, which have absorbed the same seasonality. That is information which is
real *and* duplicated, and it is a very different situation from information
that is not there. Ranking on `contribution` alone would treat the two
identically.

Some contributions come out negative. That means the model scored *better* on
average without the covariate, which on this many stations is noise rather than
a finding — and it is exactly the case the test exists to keep you from
over-reading.

```{r jackknife-test}
jk[, c("variable", "contribution", "statistic", "p_value", "p_adjusted",
       "significant")]
```

Nothing here is significant, and that is the right answer rather than a broken
test. Three covariates over a few hundred synthetic stations, scored on five
folds, is not enough evidence to establish that any one of them is load-bearing.
A test that returned confident answers from this much data would be the one to
distrust.

`p_value` is one-sided on the per-fold differences, with the variance inflated
to account for the folds sharing most of their training rows. At five folds that
inflation is a factor of 1.5 on the standard error, so every statistic here is
two-thirds of what a naive paired *t*-test would have reported. It makes no
difference to the verdict on this run — these differences are nowhere near the
boundary either way — but it is the difference between a real finding and a
false one when a covariate lands close to it, which on real data is where the
interesting ones land. `p_adjusted` then accounts for having asked once per
covariate. A GLM or a GAM would also fill in `parametric_p` with a likelihood
ratio test; a forest has no likelihood, so it is `NA` here.

**Nothing is dropped.** That is the default, and it is deliberate:

```{r jackknife-drop}
jackknife_settings(config)   # NULL: this config has no jackknife block

# What dropping *would* have removed, under the settings the test ran with.
jackknife_dropped(jk)
```

Only one name comes back even though all three failed, because
`min_predictors` floors the model at two covariates and the ones handed back are
the strongest of those that failed.

A covariate that fails this test is one the *others already account for on these
stations*, which is a statement about collinearity in this sample at least as
much as about ecology. Depth and surface temperature carry much of the same
information on a shelf, and the test will call either one redundant depending on
which the model reached for first. Turning `covariates.jackknife.drop` on tells
the pipeline to act on it anyway, and `keep` protects a covariate that is in the
model because the study is about it:

```{r jackknife-config, eval = FALSE}
config$covariates$jackknife <- list(drop = TRUE, keep = "jday", workers = 4)
result <- run_taupatch(config)

result$jackknife                      # the table, also written to the run
result$config$covariates$exclude      # what came out
```

## Fitting all the models at once

Rather than choosing an algorithm, fit several and combine them. Set
`model.type` to `ensemble` and the pipeline does the rest — everything after the
fit works the same, so nothing else in the config changes.

Four types are available; two are used here because `xgboost` and `mgcv` are
suggested rather than required.

```{r ensemble-fit, message = FALSE, warning = FALSE}
config$model$ensemble <- list(types = c("rf", "glm"), rule = "weighted_mean")
config$paths$output_dir <- file.path(tempdir(), "taupatch_vignette_ens")

result <- run_taupatch(config)
result$model
```

Each member's score, and the weight it earned, is the first thing to read. A
table showing one member near 1.0 and the rest near zero is a single model with
extra steps.

```{r ensemble-summary}
result$model$summary
```

The ensemble's evaluation is **its own**, not the average of its members'. Every
member was fitted on the same folds from the same seed, so their held-out
predictions line up row for row, and the combined out-of-fold predictions go
through the same evaluation as a single model's:

```{r ensemble-eval}
result$model$evaluation[1:2, c("metric", "value", "lower", "upper")]

# Each member's own, for comparison.
vapply(result$model$members,
       function(m) m$evaluation$value[m$evaluation$metric == "roc_auc" &
                                        is.na(m$evaluation$threshold)],
       numeric(1))
```

That distinction matters. Combining members that make *different* mistakes beats
all of them; combining members that make the same mistakes does not, and only a
cross-validated ensemble prediction can tell those apart.

Importance is weighted across members, with each member's own kept beside it — a
predictor the forest leans on and the GLM ignores is a fact about the shape of
the relationship, and the average is the one number that hides it.

```{r ensemble-importance}
result$model$importance
```

Every combination rule is computed and written, so a committee map can be read
off the same file without refitting. `rule` only picks which one becomes the
`suitability` layer:

```{r ensemble-layers}
names(terra::rast(result$projections$geotiff[1]))
```

`algorithm_sd` is the new one, and it is **not** the same as `suitability_sd`
from the section above. That one is a single algorithm refitted on resampled
stations; this one is different algorithms looking at the same stations and
disagreeing. Both can be on at once, and a cell can be quiet on one and loud on
the other:

```{r ensemble-spread}
ens <- readr::read_csv(
  file.path(config$paths$output_dir, "projections", "suitability.csv"),
  show_col_types = FALSE
)

summary(ens[, c("suitability", "algorithm_sd", "suitability_sd")])
```

## Moving to real data

Three things change, and nothing else does.

**The station file.** Point `paths.zoop_file` at your own CSV. If you have the
raw NOAA EcoMon export rather than a formatted database, reshape it first —
`format_zoop_data()` splits the date, renames the coordinate columns, strips the
units off each taxon column and divides the abundances accordingly, and carries
everything else through untouched.

```{r format, eval = FALSE}
format_zoop_data("raw_ecomon.csv", write_to = "data/zooplankton.csv")
zoop_taxa("raw_ecomon.csv")   # every taxon the file carries
```

**The covariate source.** `covariates.source: copernicus` instead of `mock`, and
the [Copernicus Marine
Toolbox](https://help.marine.copernicus.eu/en/collections/4060068-copernicus-marine-toolbox)
installed and logged in. The covariate names stay the same:

```{r catalog}
covariate_info()[1:8, c("name", "label", "units", "spatial")]
```

**The window and the area.** `dates`, `projection`, and `study_area.bbox`. Draw
the bounding box wider than your stations if you use gradients or fronts, which
are undefined on the edge of the grid.

Write the new config rather than hand-editing YAML — `generate_config()`
validates as it writes, so a mistyped covariate is an error at that call:

```{r generate, eval = FALSE}
generate_config("cfin_gom",
                zoop_file = "data/zooplankton.csv",
                selected = c("SST", "SSS", "CHL"),
                bathymetry = "DEPTH",
                transform = list(log1p = c("CHL", "DEPTH")))
```

## Doing it without writing code

The same pipeline is a GUI. Every config field is a control, the run streams its
progress, and **Download config** turns whatever you clicked into a YAML file you
can re-run or hand to someone else.

```{r app, eval = FALSE}
run_taupatch_app()
```

## Where to go next

The README covers each config block in depth — thresholds, derived covariates,
transformations, resampling before the join, the four model types and the
diagnostics each one contributes. In R:

- `?load_config`, `?generate_config` — the config
- `?covariate_info`, `?derivoce_covariates` — what can be a predictor
- `?covariate_transforms` — and what can be done to one
- `?model_types` — the four models, and what each is good for
- `?partial_effects`, `?gam_smooth_terms` — reading a fitted model
- `?jackknife_covariates`, `?jackknife_settings` — testing which covariates earn
  their place
- `?fit_patch_ensemble`, `?ensemble_settings`, `?ensemble_rules` — combining
  several model types

## Citing a run

taupatch is mostly plumbing between other people's data and other people's
methods. A run reaches Copernicus Marine products, NOAA terrain, and station data
none of which belong to this package, and fits models and metrics that are each
somebody's paper. `citation("taupatch")` gives this package and the paper it
implements; the [References section of the
README](https://github.com/chross22/taupatch#references) groups everything else
so you can pick out the subset a particular run actually used. Each function's
own help page carries the references relevant to it.

## References

taupatch is mostly plumbing between other people's data and other people's
methods, and the obligation to cite travels with those rather than with this
package. **Cite whichever of these your run actually used** — the groupings below
are meant to make that easy to work out. Each function's own `?help` carries the
references relevant to it, and `covariate_info()` names the source of every
covariate at runtime.

### The model this implements

- Ross CH, Runge JA, Roberts JJ, Brady DC, Tupper B, Record NR (2023). Estimating
  North Atlantic right whale prey based on *Calanus finmarchicus* thresholds.
  *Marine Ecology Progress Series* **703**, 1–16.
  [doi:10.3354/meps14204](https://doi.org/10.3354/meps14204)
- Elith J, Leathwick JR (2009). Species distribution models: ecological
  explanation and prediction across space and time. *Annual Review of Ecology,
  Evolution, and Systematics* **40**, 677–697.
  [doi:10.1146/annurev.ecolsys.110308.120159](https://doi.org/10.1146/annurev.ecolsys.110308.120159)
- Thuiller W, Lafourcade B, Engler R, Araújo MB (2009). BIOMOD – a platform for
  ensemble forecasting of species distributions. *Ecography* **32**(3), 369–373.
  [doi:10.1111/j.1600-0587.2008.05742.x](https://doi.org/10.1111/j.1600-0587.2008.05742.x)
  — the framework the archived `original/` pipeline was built on, replaced here by
  tidymodels

### Station data

Not distributed with this package. A run on the NOAA export should cite it:

- NOAA National Marine Fisheries Service, Northeast Fisheries Science Center.
  *Zooplankton and ichthyoplankton abundance and distribution in the North
  Atlantic collected by the Ecosystem Monitoring (EcoMon) Project*. NOAA National
  Centers for Environmental Information, NCEI Accession 0187513.
  <https://www.ncei.noaa.gov/archive/accession/0187513>

### Environmental covariates

Fetched through [datamatch](https://github.com/chross22/datamatch), whose
[reference list](https://github.com/chross22/datamatch#references) is the complete
one for the data sources. The three products this package's catalog draws on:

- **Global Ocean Physics Reanalysis** (GLORYS12V1) — `SST`, `SSS`, `BOTT`, `UO`,
  `VO`, `SSH`, `MLD`, `SIC`. E.U. Copernicus Marine Service Information.
  [doi:10.48670/moi-00021](https://doi.org/10.48670/moi-00021)
- **Global Ocean Colour** (Copernicus-GlobColour) — satellite `CHL`, `PP`,
  `DIATO`, `DINO`. E.U. Copernicus Marine Service Information.
  [doi:10.48670/moi-00281](https://doi.org/10.48670/moi-00281)
- **Global Ocean Biogeochemistry Hindcast** — `NO3`, `PO4`, `O2`, `PH`,
  `CHL_MODEL`, `NPP_MODEL`. E.U. Copernicus Marine Service Information.
  [doi:10.48670/moi-00019](https://doi.org/10.48670/moi-00019)
- E.U. Copernicus Marine Service. *Copernicus Marine Toolbox*
  (`copernicusmarine`). <https://toolbox-docs.marine.copernicus.eu/> — how the
  above are downloaded. Mercator Ocean International publishes no DOI for the
  toolbox itself, so credit it by name; the citation that matters is the
  product's.

Seafloor terrain, for `covariates.bathymetry`:

- NOAA National Centers for Environmental Information (2022). *ETOPO 2022
  15 Arc-Second Global Relief Model*.
  [doi:10.25921/fd45-gt74](https://doi.org/10.25921/fd45-gt74)
- Pante E, Simon-Bouhet B (2013). marmap: A package for importing, plotting and
  analyzing bathymetric and topographic data in R. *PLoS ONE* **8**(9), e73051.
  [doi:10.1371/journal.pone.0073051](https://doi.org/10.1371/journal.pone.0073051)

Climate indices, when `covariates.climate` is set, come from
[datamatch](https://github.com/chross22/datamatch#climate-indices); two of the six
(`LCR`, `AMOC`) are the published output of specific work and are cited there.

### Derived covariates

Computed by [derivoce](https://github.com/chross22/derivoce), whose
[reference list](https://github.com/chross22/derivoce#references) is the complete
one. The methods behind the steps this README shows:

- Belkin IM, O'Reilly JE (2009). An algorithm for oceanic front detection in
  chlorophyll and SST satellite imagery. *Journal of Marine Systems* **78**(3),
  319–326.
  [doi:10.1016/j.jmarsys.2008.11.018](https://doi.org/10.1016/j.jmarsys.2008.11.018)
  — `distance_to_front`
- d'Ovidio F, Fernández V, Hernández-García E, López C (2004). Mixing structures
  in the Mediterranean Sea from finite-size Lyapunov exponents. *Geophysical
  Research Letters* **31**(17).
  [doi:10.1029/2004GL020328](https://doi.org/10.1029/2004GL020328) — `fsle`
- Haller G (2015). Lagrangian coherent structures. *Annual Review of Fluid
  Mechanics* **47**, 137–162.
  [doi:10.1146/annurev-fluid-010313-141322](https://doi.org/10.1146/annurev-fluid-010313-141322)
  — `ftle`, `fsle`

Coastlines for `distance_to_shore` are [Natural
Earth](https://www.naturalearthdata.com/), public domain, via `rnaturalearth`.

### Transformations

- Box GEP, Cox DR (1964). An analysis of transformations. *Journal of the Royal
  Statistical Society: Series B* **26**(2), 211–252.
  [doi:10.1111/j.2517-6161.1964.tb00553.x](https://doi.org/10.1111/j.2517-6161.1964.tb00553.x)
  — `boxcox`
- Field JG, Clarke KR, Warwick RM (1982). A practical strategy for analysing
  multispecies distribution patterns. *Marine Ecology Progress Series* **8**,
  37–52. [doi:10.3354/meps008037](https://doi.org/10.3354/meps008037) — the fourth
  root as the plankton standard
- Yeo I-K, Johnson RA (2000). A new family of power transformations to improve
  normality or symmetry. *Biometrika* **87**(4), 954–959.
  [doi:10.1093/biomet/87.4.954](https://doi.org/10.1093/biomet/87.4.954) —
  `yeojohnson`

### Models

- Araújo MB, New M (2007). Ensemble forecasting of species distributions.
  *Trends in Ecology & Evolution* **22**(1), 42–47.
  [doi:10.1016/j.tree.2006.09.010](https://doi.org/10.1016/j.tree.2006.09.010) —
  why an ensemble of algorithms rather than a chosen best one
- Breiman L (2001). Random forests. *Machine Learning* **45**(1), 5–32.
  [doi:10.1023/A:1010933404324](https://doi.org/10.1023/A:1010933404324) — `rf`,
  and the origin of permutation importance
- Chen T, Guestrin C (2016). XGBoost: a scalable tree boosting system.
  *Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge
  Discovery and Data Mining*, 785–794.
  [doi:10.1145/2939672.2939785](https://doi.org/10.1145/2939672.2939785) — the
  `brt` engine
- Elith J, Leathwick JR, Hastie T (2008). A working guide to boosted regression
  trees. *Journal of Animal Ecology* **77**(4), 802–813.
  [doi:10.1111/j.1365-2656.2008.01390.x](https://doi.org/10.1111/j.1365-2656.2008.01390.x)
- Friedman JH (2001). Greedy function approximation: a gradient boosting machine.
  *Annals of Statistics* **29**(5), 1189–1232.
  [doi:10.1214/aos/1013203451](https://doi.org/10.1214/aos/1013203451) — `brt`, and
  the partial dependence plot
- Hastie T, Tibshirani R (1986). Generalized additive models. *Statistical
  Science* **1**(3), 297–310.
  [doi:10.1214/ss/1177013604](https://doi.org/10.1214/ss/1177013604) — `gam`
- Marmion M, Parviainen M, Luoto M, Heikkinen RK, Thuiller W (2009). Evaluation
  of consensus methods in predictive species distribution modelling. *Diversity
  and Distributions* **15**(1), 59–69.
  [doi:10.1111/j.1472-4642.2008.00491.x](https://doi.org/10.1111/j.1472-4642.2008.00491.x)
  — the ensemble combination rules, compared against each other
- Marra G, Wood SN (2011). Practical variable selection for generalized additive
  models. *Computational Statistics & Data Analysis* **55**(7), 2372–2387.
  [doi:10.1016/j.csda.2011.02.004](https://doi.org/10.1016/j.csda.2011.02.004) —
  `select_features` and the `ts` basis
- McCullagh P, Nelder JA (1989). *Generalized Linear Models*, 2nd edition. Chapman
  and Hall. [doi:10.1007/978-1-4899-3242-6](https://doi.org/10.1007/978-1-4899-3242-6)
  — `glm`
- Wood SN (2011). Fast stable restricted maximum likelihood and marginal
  likelihood estimation of semiparametric generalized linear models. *Journal of
  the Royal Statistical Society: Series B* **73**(1), 3–36.
  [doi:10.1111/j.1467-9868.2010.00749.x](https://doi.org/10.1111/j.1467-9868.2010.00749.x)
  — `model.method`
- Wood SN (2017). *Generalized Additive Models: An Introduction with R*, 2nd
  edition. Chapman and Hall/CRC.
  [doi:10.1201/9781315370279](https://doi.org/10.1201/9781315370279) — the `mgcv`
  reference
- Wright MN, Ziegler A (2017). ranger: a fast implementation of random forests for
  high dimensional data in C++ and R. *Journal of Statistical Software* **77**(1),
  1–17. [doi:10.18637/jss.v077.i01](https://doi.org/10.18637/jss.v077.i01) — the
  `rf` engine

### Evaluation and diagnostics

- Allouche O, Tsoar A, Kadmon R (2006). Assessing the accuracy of species
  distribution models: prevalence, kappa and the true skill statistic (TSS).
  *Journal of Applied Ecology* **43**(6), 1223–1232.
  [doi:10.1111/j.1365-2664.2006.01214.x](https://doi.org/10.1111/j.1365-2664.2006.01214.x)
  — `tss`, and the cutoff that maximises it
- Bengio Y, Grandvalet Y (2004). No unbiased estimator of the variance of k-fold
  cross-validation. *Journal of Machine Learning Research* **5**, 1089–1105.
  <https://jmlr.org/papers/v5/grandvalet04a.html> — why the jackknife's test
  needs a variance correction rather than a better estimator
- Bouckaert RR, Frank E (2004). Evaluating the replicability of significance
  tests for comparing learning algorithms. *Advances in Knowledge Discovery and
  Data Mining* (PAKDD 2004), Lecture Notes in Computer Science **3056**, 3–12.
  Springer. — the correction applied to k-fold specifically. Listed without its
  DOI on purpose: a Springer chapter identifier contains an underscore, which
  the citation checker's DOI pattern reads as the end of the identifier
- Dietterich TG (1998). Approximate statistical tests for comparing supervised
  classification learning algorithms. *Neural Computation* **10**(7), 1895–1923.
  [doi:10.1162/089976698300017197](https://doi.org/10.1162/089976698300017197) —
  the inflated Type I error of the uncorrected paired test
- Elith J, Phillips SJ, Hastie T, Dudík M, Chee YE, Yates CJ (2011). A
  statistical explanation of MaxEnt for ecologists. *Diversity and
  Distributions* **17**(1), 43–57.
  [doi:10.1111/j.1472-4642.2010.00725.x](https://doi.org/10.1111/j.1472-4642.2010.00725.x)
  — the leave-one-out / only-one jackknife pair
- Nadeau C, Bengio Y (2003). Inference for the generalization error. *Machine
  Learning* **52**(3), 239–281.
  [doi:10.1023/A:1024068626366](https://doi.org/10.1023/A:1024068626366) — the
  variance correction behind the jackknife's `p_value` and `compare_runs()`
- Hijmans RJ (2012). Cross-validation of species distribution models: removing
  spatial sorting bias and calibration with a null model. *Ecology* **93**(3),
  679–688. [doi:10.1890/11-0826.1](https://doi.org/10.1890/11-0826.1) — the
  `ssb` row in `evals.csv`
- Schoener TW (1968). The *Anolis* lizards of Bimini: resource partitioning in
  a complex fauna. *Ecology* **49**(4), 704–726.
  [doi:10.2307/1935534](https://doi.org/10.2307/1935534) — Schoener's *D* in
  `projection_overlap()`
- Warren DL, Glor RE, Turelli M (2008). Environmental niche equivalency versus
  conservatism: quantitative approaches to niche evolution. *Evolution*
  **62**(11), 2868–2883.
  [doi:10.1111/j.1558-5646.2008.00482.x](https://doi.org/10.1111/j.1558-5646.2008.00482.x)
  — Warren's *I*
- Hoenig JM, Heisey DM (2001). The abuse of power: the pervasive fallacy of
  power calculations for data analysis. *The American Statistician* **55**(1),
  19–24.
  [doi:10.1198/000313001300339897](https://doi.org/10.1198/000313001300339897)
  — why `compare_runs()` reports a minimum detectable difference rather than
  the “observed power” it is often confused with
- Fisher A, Rudin C, Dominici F (2019). All models are wrong, but many are useful:
  learning a variable's importance by studying an entire class of prediction
  models simultaneously. *Journal of Machine Learning Research* **20**(177), 1–81.
  <https://jmlr.org/papers/v20/18-760.html> — permutation importance as a
  model-agnostic quantity, which is what makes it comparable across the four types
- Saito T, Rehmsmeier M (2015). The precision-recall plot is more informative than
  the ROC plot when evaluating binary classifiers on imbalanced datasets. *PLoS
  ONE* **10**(3), e0118432.
  [doi:10.1371/journal.pone.0118432](https://doi.org/10.1371/journal.pone.0118432)
  — `pr_auc`
- Elith J, Kearney M, Phillips S (2010). The art of modelling range-shifting
  species. *Methods in Ecology and Evolution* **1**(4), 330–342.
  [doi:10.1111/j.2041-210X.2010.00036.x](https://doi.org/10.1111/j.2041-210X.2010.00036.x)
  — the multivariate environmental similarity surface behind `novelty`
- Sofaer HR, Hoeting JA, Jarnevich CS (2019). The area under the precision-recall
  curve as a performance metric for rare binary events. *Methods in Ecology and
  Evolution* **10**(4), 565–577.
  [doi:10.1111/2041-210X.13140](https://doi.org/10.1111/2041-210X.13140)

### Software this is built on

- Chang W, Cheng J, Allaire JJ, Sievert C, Schloerke B, Xie Y, Allen J, McPherson
  J, Dipert A, Borges B. *shiny: Web Application Framework for R*.
  [doi:10.32614/CRAN.package.shiny](https://doi.org/10.32614/CRAN.package.shiny) —
  `run_taupatch_app()`, with `leaflet` and `shinyFiles`
- Garnier S, Ross N, Rudis R, Camargo AP, Sciaini M, Scherer C. *viridisLite:
  Colorblind-Friendly Color Maps (Lite Version)*.
  [doi:10.32614/CRAN.package.viridisLite](https://doi.org/10.32614/CRAN.package.viridisLite)
  — the magma scale used throughout, from the colormaps designed for matplotlib by
  Stéfan van der Walt and Nathaniel Smith
- Hijmans RJ. *terra: Spatial Data Analysis*.
  [doi:10.32614/CRAN.package.terra](https://doi.org/10.32614/CRAN.package.terra) —
  the projection rasters
- Kuhn M, Wickham H (2020). *Tidymodels: a collection of packages for modeling and
  machine learning using tidyverse principles*. <https://www.tidymodels.org> —
  `parsnip`, `recipes`, `rsample`, `tune`, `workflows`, `yardstick`
- Pebesma E (2018). Simple features for R: standardized support for spatial vector
  data. *The R Journal* **10**(1), 439–446.
  [doi:10.32614/RJ-2018-009](https://doi.org/10.32614/RJ-2018-009) — `sf`
- Wickham H (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer.
  [doi:10.1007/978-3-319-24277-4](https://doi.org/10.1007/978-3-319-24277-4)

`citation()` works on any of the R packages above.
